---
title: "01 Import and Summarize"
output: rmarkdown::html_vignette
description: |
  tidyproteomics: simplified data object and analysis functions
vignette: >
  %\VignetteIndexEntry{Workflows: Simple}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1041)
options(dplyr.print_max = 10)
```

The following is a simple workflow for the importing and summarizing of a data set. The tidyproteomics data object has a defined `print()` function that will summarized the data contents, while the `summary()` function will provide a statistical summary of the quantitative data with values as described in Table 1.

*Table 1 - summary table description*

| Colunm          | Accounting                               |
|-----------------|------------------------------------------|
| *first*         | group (eg sample name)                   |
| files           | *integer* number present in group        |
| proteins        | *integer* number present in group        |
| protein_groups  | *integer* number present in group        |
| peptides        | *integer* number present in group        |
| peptides_unique | *integer* number present in group        |
| quantifiable    | *percent* ratio of non-zero values       |
| CVs             | *ratio* quantitative abundance (sd/mean) |

```{r, message = FALSE}
library("dplyr")
library("tidyproteomics")
```

### Importing Data

```{r}
# path_to_package_data() loads data specific to this package
# for your project load local data
# example: 
# your_proteins <- "./data/your_exported_results.xlsx" %>%
#   import("ProteomeDiscoverer", "proteins")

ecoli_proteins <- path_to_package_data("pd_ecoli-hint_proteins.xlsx") %>%
  import("ProteomeDiscoverer", "proteins")
```

#### Print Data Contents

Printing the imported data object, or simply exposing the object will show a summary of the data object contents

```{r}
ecoli_proteins
```

As more operations are performed on the data, more of the contents are summarized

```{r}
ecoli_proteins %>%
  reassign(field = 'sample', pattern = 'hint_ko', replace = 'ko') %>%
  expression(ko/wt) %>%
  enrichment(ko/wt, .term = 'biological_process') %>%
  enrichment(ko/wt, .term = 'molecular_function')
```

#### Summarize Quantitative Data

Use the explicit `summary()` function summarize the data, in this case globally.

```{r}
ecoli_proteins %>% summary()
```

Here is a summary by unique sample names

```{r}
ecoli_proteins %>% summary(by = 'sample') 
```

A summary that includes contamination where the description is contains 'CRAP', as in the crap-ome

```{r}
ecoli_proteins %>% summary(contamination = 'CRAP') 
```

A summary for contamination where we specify where the description is contains 'ribosome'

```{r}
ecoli_proteins %>% summary(contamination = "ribosome") 
```

A summary based on a term set in the provided annotations

```{r}
ecoli_proteins %>% summary('biological_process')
```
