---
title: "Enrichment"
output: rmarkdown::html_vignette
description: |
  tidyproteomics: simplified data object and analysis functions
vignette: >
  %\VignetteIndexEntry{Subsetting}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1041)
options(dplyr.print_max = 10)
```

### Term Enrichment

Term enrichment analysis is possible proceeding an expression analysis with the `enrichment()` function, employing either the _GSEA_ algorithm or a simple _Wilcoxon rank sum_ comparison, and can be visualized with the `plot_enrichment()` function.

```{r, message=FALSE, warning=FALSE}
library("dplyr")
library("tidyproteomics")

rdata <- ecoli_proteins %>% 
  normalize(.method = 'linear') %>%
  impute(impute_function = base::min) %>%
  expression(ko/wt) %>%
  enrichment(ko/wt, .term = 'biological_process')
```

#### Exporting Results
The results of this analysis are stored in the tidyproteomics data-object and can easily be exported to save as a flat data table or used in an external down stream analysis.
```{r}

rdata %>% export_analysis(ko/wt, .analysis = 'enrichment', .term = 'biological_process')
```

#### Enrichment Plot

```{r}

rdata %>% plot_enrichment(ko/wt, .term = 'biological_process')
```
