---
title: "Enrichment"
output: rmarkdown::html_vignette
description: |
tidyproteomics: simplified data object and analysis functions
vignette: >
  %\VignetteIndexEntry{Enrichment}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1041)
options(dplyr.print_max = 10)
```

### Term Enrichment

#### Available Methods

+---------------+------------------------------------------------------------------------------------------+
| GSEA          | Korotkevich G, Sukhov V, Sergushichev A (2019).                                          |
|               | “Fast gene set enrichment analysis.”                                                     |
|               | bioRxiv. doi:10.1101/060012,                                                             |
|               | http://biorxiv.org/content/early/2016/06/20/060012.                                      |
|               |                                                                                          |
|               | [Bioconductor::fgsea](https://bioconductor.org/packages/release/bioc/html/fgsea.html)    |
+---------------+------------------------------------------------------------------------------------------+
| Wilcoxon      | H. B. Mann, D. R. Whitney "On a Test of Whether one of Two Random Variables is           |
| Rank Sum      | Stochastically Larger than the Other," Ann. Math. Statist. 18(1), 50-60, (March, 1947)   |
|               |                                                                                          |
|               | [R::stats::wilcoxon]                                                                     |
+---------------+------------------------------------------------------------------------------------------+
| Fisher's      | Fisher, R. A. “On the Interpretation of χ2 from Contingency Tables, and the Calculation  |
| Excact Test   | of P.” Journal of the Royal Statistical Society, vol. 85, no. 1, 1922, pp. 87–94.        |
|               |                                                                                          |
|               | [R::stats::fisher.test]                                                                  |
+---------------+------------------------------------------------------------------------------------------+

Term enrichment analysis is possible proceeding an expression analysis with the `enrichment()` function, employing either the _GSEA_ algorithm, a simple _Wilcoxon rank sum_ comparison, or _Fisher's Exact Test_ and can be visualized with the `plot_enrichment()` function. Both _GSEA_ and _Wilcoxon_ will consider genes of a given term significant if those genes are skewed mostly up or down. The _Fisher's Exact Test_ will consider genes of a given term significant if they are "represented" more often than expected -- this is the basis of __Over-Representation Analysis__ and will consider the gene-set regardless of being skewed mostly up or down.

```{r, message=FALSE, warning=FALSE}
library("dplyr")
library("tidyproteomics")

rdata <- hela_proteins %>% 
  normalize(.method = 'linear') %>%
  expression(knockdown/control) %>%
  enrichment(knockdown/control, 
             .terms = 'biological_process', 
             .method = 'wilcoxon')
```

#### Exporting Results
The results of this analysis are stored in the tidyproteomics data-object and can easily be exported to save as a flat data table or used in an external down stream analysis.
```{r}
rdata %>% 
  export_analysis(knockdown/control, 
                  .analysis = 'enrichment', 
                  .term = 'biological_process')

```

The results can also have any other terms appended to the output, such as protein or gene_name, which may be useful for follow-on analyses.
```{r}
rdata %>% 
  export_analysis(knockdown/control, 
                  .analysis = 'enrichment', 
                  .term = 'biological_process',
                  .append = 'gene_name')

```

#### Enrichment Plot

```{r}
rdata %>% plot_enrichment(knockdown/control, .term = 'biological_process')

```
