---
title: "Expression"
output: rmarkdown::html_vignette
description: |
  tidyproteomics: simplified data object and analysis functions
vignette: >
  %\VignetteIndexEntry{Subsetting}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1041)
options(dplyr.print_max = 10)
```

### Analysis of Two Samples

One of the most fundamental analyses for quantitative proteomics is the
estimation of univariate differences between two samples. Tidyproteomics
accommodates two sample analysis using the `expression()` function, the
results of which can be visualized using the `plot_volcano()` and
`plot_proportion()` visualization functions as shown below. The
expression analysis can be accomplished thought the R *limma* package,
or using any two-sample statistical comparison, the default method
employs the two-sided T-test. If no values pass the cutoff it will
automatically choose the top proteins to display.

```{r, message = FALSE}

```

| parameter           | description                                                               | inputs                                                | default        |
|---------------------|---------------------------------------------------------------------------|-------------------------------------------------------|----------------|
| log2fc_min          | the minimum cutoff for displaying                                         | *0 to Inf*                                            | *2*            |
| significance_column | the column to use for significance                                        | *adj_p\_value, p_value*                               | *adj_p\_value* |
| significance_max    | the minimum cutoff for displaying                                         | *1 to 1/Inf*                                          | *0.05*         |
| labels_column       | how to label the points                                                   | any value in annotations, eg *gene_name, description* | *protein*      |
| show_pannels        | show the colored areas highlighting significance                          | *TRUE / FALSE*                                        | *TRUE*         |
| show_lines          | show the dotted cutoff lines                                              | *TRUE / FALSE*                                        | *TRUE*         |
| show_fc_scale       | show the second scale on top converting *log2_foldchange* to *foldchange* | *TRUE / FALSE*                                        | *TRUE*         |
| point_size          | size of each dot                                                          | *1/10* to *5*                                         | *NULL*         |
| color_positive      | color for positive expression                                             | [any R accepts](https://r-charts.com/colors/)         | *dodgerblue*   |
| color_negative      | color for negative expression                                             | [any R accepts](https://r-charts.com/colors/)         | *firebrick1*   |

```{r, message=FALSE}
library("dplyr")
library("tidyproteomics")

rdata <- ecoli_proteins %>% 
  normalize(.method = 'linear') %>%
  impute(impute_function = base::min) %>%
  expression(ko/wt)
```

#### Exporting Results

The results of this analysis are stored in the tidyproteomics
data-object and can easily be exported to save as a flat data table or
used in an external down stream analysis.

```{r}

rdata %>% export_analysis(ko/wt, .analysis = 'expression')
```

#### Volcano Plot

The default is to plot accoring to adjusted p-value with a Log2
fold-change cutoff of 2 or greater.

```{r}

rdata %>% plot_volcano(ko/wt)
```

The parameters can be adjusted to suit your needs ...

```{r}

rdata %>% plot_volcano(ko/wt,
                       significance_column = 'p_value',
                       log2fc_min = 1)
```

... or desires.

```{r}
library(ggplot2)

rdata %>% plot_volcano(ko/wt,
                       significance_column = 'p_value',
                       log2fc_min = 1,
                       color_positive = 'orange',
                       color_negative = 'purple',
                       show_lines = FALSE,
                       show_fc_scale = FALSE,
                       show_pannels = FALSE,
                       labels_column = 'gene_name') +
  labs(title = "A nice volcano plot", subtitle = "with great colors")
```

#### Proportional Plot

The proportional plot is complementary to the volcano plot and utilizes
all the same underlying data. The default is to show both the top 1%
(`proportion_min = 0.01`) along with the values passing the fold-change
and significance cutoffs.

```{r, message=FALSE}
library(ggplot2)

rdata %>% plot_proportion(ko/wt)
```

This plot can likewise also be manipulated like the volcano plot.

```{r}
library(ggplot2)

rdata %>% plot_proportion(ko/wt,
                       significance_column = 'p_value',
                       log2fc_min = 5,
                       proportion_min = 0.03,
                       color_positive = 'orange',
                       color_negative = 'purple',
                       show_lines = TRUE,
                       labels_column = 'gene_name')

```
