---
title: "Imputing"
output: rmarkdown::html_vignette
description: |
  tidyproteomics: simplified data object and analysis functions
vignette: >
  %\VignetteIndexEntry{Workflows: Simple}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1041)
options(dplyr.print_max = 10)
```

Along with normalization, imputing missing values is another important task in quantitative proteomics that can be challenging to implement given the desired method. Again, tidyproteomics attempts to facilitate this with the `impute()` function, which currently can support any base level or user defined function along with implementing the R package *missForest*, widely regarded as one of the best algorithms for missing value imputation. Note that this method is a **between** sample imputation.

While random forest algorithms have shown superiority in imputation and regression, that does not portend their use in every case. For example, imputing missing values from a knock-out experiment, such as the dataset included in this package, are preferrable to minimum value imputation over the more complex random forest, simply because in this experiment we know how the experiment is affected.

Imputation currently supports the following methods calculated either **within** or **between** samples:

| Function                 | Method                    | Description                                                        |
|---------------|---------------|------------------------------------------|
| base::min                | **within** or **between** | the minimum value in any given set                                 |
| stats::median            | **within** or **between** | the minimum value in any given set                                 |
| *user supplied function* | **within** or **between** | e.g. *function (x, na.rm) { quantile(x, 0.05, na.rm = na.rm)[1] }* |
| "randomforest"           | **between** only          | a non-linear random forest implementation of missForest            |

```{r, message = FALSE}
library("dplyr")
library("ggplot2")
library("tidyproteomics")
```

### Imputing

```{r, message=FALSE}
# path_to_package_data() loads data specific to this package
# for your project load local data
# example: 
# your_proteins <- "./data/your_exported_results.xlsx" %>%
#   import("ProteomeDiscoverer", "proteins")

rdata <- path_to_package_data("pd_ecoli-hint_proteins.xlsx") %>%
  import("ProteomeDiscoverer", "proteins")
```

#### Using *within*

Note the difference using **within** ..

```{r, message=FALSE}

rdata %>% 
  impute(impute_function = base::min, method = 'within') %>%
  subset(protein %like% "P05706|P0A8R4|P0A9K1|P0AAJ8|P0ACE7|P18776|P24252|P31826|P32674") %>% 
  extract() %>%
  ggplot(aes(replicate, abundance)) +
  geom_point(aes(color=sample)) +
  facet_wrap(~identifier) +
  scale_y_log10()
```

#### Using *between*

.. as opposed to **between**. The **between** method can be considered to contain the bias of any *real* offset, note our protein *P0ACE7* (i.e *hinT*), shows the expected offset for the **within** method, and does not for the **between** method. Consider only using **between** methods when imputing values you suspect are missing-at-random. In our case *P0ACE7* (i.e *hinT*) is missing-not-at-random, because we performed a genetic knockout mutation.

```{r, message=FALSE}

rdata %>% 
  impute(impute_function = base::min, method = 'between') %>%
  subset(protein %like% "P05706|P0A8R4|P0A9K1|P0AAJ8|P0ACE7|P18776|P24252|P31826|P32674") %>% 
  extract() %>%
  ggplot(aes(replicate, abundance)) +
  geom_point(aes(color=sample)) +
  facet_wrap(~identifier) +
  scale_y_log10()
```
