---
title: "Normalizing"
output: rmarkdown::html_vignette
description: |
  tidyproteomics: simplified data object and analysis functions
vignette: >
  %\VignetteIndexEntry{Normalizing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1041)
options(dplyr.print_max = 10)
```

Quantitative proteomics relies on accurate normalization, for which several choices are available, yet remain somewhat difficult to accurately implement or require formatting data differently for each. For example, a simple alignment of measured median's requires only a few lines of code, while implementing normalization from the ***limma*** package requires non-intuitive formatting of the data. The `normalize()` function is designed as a wrapper to handle various methods of normalization all at once, then allowing researchers the ability to examine the result and choose the method best suited for their analysis. Alternatively, the `select_normalization()` function can automatically select the best normalization based on a weighted score combining CVs, dynamic range and variability in the first three PCA components, or the user can override this selection manually. The values from the selected normalization are then use for all downstream plots and analyses such as `expression()` and `enrichment()`.

Normalization currently supports the following methods:

| Method       | Description                                                                                                                                                                                                                  |
|------------------|------------------------------------------------------|
| scaled       | values are aligned medians and dynamic ranges to adjusted to be similar                                                                                                                                                      |
| median       | values are simply aligned to the same median value                                                                                                                                                                           |
| linear       | a linear regression is applied to each sample according to individual analyte means                                                                                                                                          |
| limma        | wrapper to the limma package                                                                                                                                                                                                 |
| loess        | a non-linear loess (quantile) regression applied to each sample according to individual analyte means                                                                                                                        |
| svm          | a non-linear regression (support vector machine) applied to each sample according to individual analyte means. This implementation utilizes eps-regression from the e1071 package and tuning of the gama and cost parameters. |
| randomforest | a non-linear random forest regression applied to each sample according to individual analyte means                                                                                                                           |

```{r, message = FALSE}
library("dplyr")
library("tidyproteomics")
```

### Normalization of the whole proteome

```{r, message=FALSE}
# path_to_package_data() loads data specific to this package
# for your project load local data
# example: 
# your_proteins <- "./data/your_exported_results.xlsx" %>%
#   import("ProteomeDiscoverer", "proteins")

rdata <- path_to_package_data("p97KD_HCT116") %>%
  import("ProteomeDiscoverer", "proteins") %>%
  normalize()
```

### Visualizing

Plotting the overall variation and dynamic range. It is easy to reduce variability simply by squashing the dynamic range - plotting that here can ensure that does not happen. Keep in mind that the overall dynamic range for several mis-aligned sets will be smaller when properly normalized - often *median* normalization has an immediate impact on the CVs and is the least rigorous. Note here the dramatic reduction in CVs *randomforest* yielded while also maintaining the overall dynamic range.

#### CVs and Dynamic Range

```{r}
rdata %>% plot_variation_cv()
```

#### PCA

```{r}
rdata %>% plot_variation_pca()
```

#### Effect of Normalization on Dynamic Range

The effect *randomforest* has on the data is most noticable at the lower dynamic range, as seen in the plot below where the CVs for values under 10\^6 in abundance have been dramatically reduced. If you are hunting for low abundant biomarkers, this method is highly recomended.

```{r, message=FALSE}
path_to_package_data("p97KD_HCT116") %>%
  import("ProteomeDiscoverer", "proteins") %>%
  normalize(.method = c('median', 'randomforest')) %>% 
  plot_dynamic_range()
```

### Normalization based on a subset

In addition to proteome wide normalization, a subset can be used as the basis for normalization, such as the case for spike-in quantitative analytes or the perhaps the bait protein in an immunoprecipitation experiment. This is accomplished with the same semantic syntax as with the `subset()` function and is reflected in the recorded operations. Be aware that using a regression from a smaller dynamic range than the full proteome may not predict well beyond the limts of the subset. 

NOTE: while *linear* and *svm* can predict beyond the training set limits, *loess* can not.

NOTE: The *randomforest* normalization method, as implemented, requires the the prediction set size to match the training test size, and therefore is only applicable to normalization of the whole proteome. The *limma* normalization method also does not accept a subset for normalization. Both of these methods will be skipped if a subset is implemented.

```{r, message=FALSE}
rdata <- path_to_package_data("p97KD_HCT116") %>%
  import("ProteomeDiscoverer", "proteins") %>%
  normalize(description %like% 'ribosome',
            .method = c('median', 'linear', 'loess', 'svm')) 

rdata %>% plot_variation_cv()
```

```{r}
rdata %>% plot_dynamic_range()
```

```{r}
rdata %>% operations()
```
