---
title: "Normalizing"
output: rmarkdown::html_vignette
description: |
  tidyproteomics: simplified data object and analysis functions
vignette: >
  %\VignetteIndexEntry{Normalizing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1041)
options(dplyr.print_max = 10)
```

Quantitative proteomics relies on accurate normalization, for which several choices are available, yet remain somewhat difficult to accurately implement or require formatting data differently for each. For example, a simple alignment of measured median's requires only a few lines of code, while implementing normalization from the ***limma*** package requires non-intuitive formatting of the data. The `normalize()` function is designed as a wrapper to handle various methods of normalization all at once, then allowing researchers the ability to examine the result and choose the method best suited for their analysis. Alternatively, the `select_normalization()` function can automatically select the best normalization based on a weighted score combining CVs, dynamic range and variability in the first three PCA components, or the user can override this selection manually. The values from the selected normalization are then use for all downstream plots and analyses such as `expression()` and `enrichment()`.

Normalization in tidyproteomics attempts to apply each function universally, meaning the same towards peptide and protein values. To do this each data-object contains a variable called the `identifier` , this tells the underlying helper functions what values in the quantitative table "identify" the thing being measured.

```{r, echo=FALSE}
library(tidyproteomics)
library(tidyverse)
```
Proteins have a single identifier ...
```{r}
hela_proteins$identifier
hela_proteins$quantitative %>% head()
```

Peptides have multiple identifiers ...
```{r}
hela_peptides$identifier
hela_peptides$quantitative %>% head()
```

### Normalization Functions

Normalization currently supports the methods described below.

> *Benchmark timing was accomplished on an M1 MacBook Pro (macOS Ventura 13.3 22E252). Parallel processing for both `randomforest` and `svm` are accomplished using `mclapply` from the R parallel package.*

+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------+
| Method       | Description                                                                                                                                                                                                                   | Timing                    |
+==============+===============================================================================================================================================================================================================================+===========================+
| scaled       | values are aligned medians and dynamic ranges to adjusted to be similar                                                                                                                                                       | 0.3 sec                   |
+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------+
| median       | values are simply aligned to the same median value                                                                                                                                                                            | 0.3 sec                   |
+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------+
| linear       | a linear regression is applied to each sample according tosqr individual analyte means                                                                                                                                        | 0.2 sec                   |
+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------+
| limma        | wrapper to the limma package                                                                                                                                                                                                  | 0.4 sec                   |
+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------+
| loess        | a non-linear loess (quantile) regression applied to each sample according to individual analyte means                                                                                                                         | 1.3 sec                   |
+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------+
| svm          | a non-linear regression (support vector machine) applied to each sample according to individual analyte means. This implementation utilizes eps-regression from the e1071 package and tuning of the gama and cost parameters. | 5min & 52.1 sec           |
|              |                                                                                                                                                                                                                               |                           |
|              |                                                                                                                                                                                                                               | 2min & 26.5 sec (4 cores) |
+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------+
| randomforest | a non-linear random forest regression applied to each sample according to individual analyte means.                                                                                                                           | 37.8 sec                  |
|              |                                                                                                                                                                                                                               |                           |
|              |                                                                                                                                                                                                                               | 14.0 sec (4 cores)        |
+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------+

```{r, message = FALSE}
library("dplyr")
library("tidyproteomics")
```

### Normalization of the whole proteome

```{r, message=FALSE}
# path_to_package_data() loads data specific to this package
# for your project load local data
# example: 
# your_proteins <- "./data/your_exported_results.xlsx" %>%
#   import("ProteomeDiscoverer", "proteins")

rdata <- hela_proteins %>% normalize(.method = c("scaled", "median", "linear", "loess", "randomforest"))
```

### Visualizing

Plotting the overall variation and dynamic range. It is easy to reduce variability simply by squashing the dynamic range - plotting that here can ensure that does not happen. Keep in mind that the overall dynamic range for several mis-aligned sets will be smaller when properly normalized - often *median* normalization has an immediate impact on the CVs and is the least rigorous. Note here the dramatic reduction in CVs *randomforest* yielded while also maintaining the overall dynamic range.

#### CVs and Dynamic Range

```{r}
rdata %>% plot_variation_cv()
```

#### PCA

```{r}
rdata %>% plot_variation_pca()
```

#### Effect of Normalization on Dynamic Range

The effect *randomforest* has on the data is most noticable at the lower dynamic range, as seen in the plot below where the CVs for values under 10\^6 in abundance have been dramatically reduced. If you are hunting for low abundant biomarkers, this method is highly recomended.

```{r, message=FALSE}
hela_proteins %>%
  normalize(.method = c('median', 'randomforest')) %>% 
  plot_dynamic_range()
```

### Normalization based on a subset

In addition to proteome wide normalization, a subset can be used as the basis for normalization, such as the case for spike-in quantitative analytes or the perhaps the bait protein in an immunoprecipitation experiment. This is accomplished with the same semantic syntax as with the `subset()` function and is reflected in the recorded operations. Be aware that using a regression from a smaller dynamic range than the full proteome may not predict well beyond the limts of the subset.

NOTE: while *linear* and *svm* can predict beyond the training set limits, *loess* can not.

NOTE: The *randomforest* normalization method, as implemented, requires the the prediction set size to match the training test size, and therefore is only applicable to normalization of the whole proteome. The *limma* normalization method also does not accept a subset for normalization. Both of these methods will be skipped if a subset is implemented.

```{r, message=FALSE}
rdata <- hela_proteins %>%
  normalize(description %like% 'ribosome',
            .method = c('median', 'linear', 'loess', 'svm')) 

rdata %>% plot_variation_cv()
```

```{r}
rdata %>% plot_dynamic_range()
```

```{r}
rdata %>% operations()
```

### Developing Additional Methods

The normalization function can accommodate new methods fairly easily, but not without being tested and integrated with the package as a whole. This is due to the serial deployment of each normalization method and subsequent selection of the best method going forward. That could change in the future. For now lets walk through the process of adding a new method. And as always, reach out the the maintainers to have your method added to the package for others to use.

#### Where the normalization gets called

In the code base locate the file `R > control_normalization.R`, this is where the method gets applied. Note, near the bottom, the `if()` statements, specifically for `median`, this is where median normalization gets implemented.

``` r
## d:        the raw data for all samples
## dc_this:  the 'centered' data
## d_norm:   the output normalized data

if(m == 'median') {         d_norm <- d %>% normalize_median(dc_this)}
```

All normalization functions require two inputs, a table of all the raw values log2 transformed, and a table of grouped and 'centered' values to normalize against. This table can either be computed from the entire proteome or from a subset as demonstrated above. 

``` r
## the log2 raw values

# A tibble: 30,157 × 4
   sample    replicate identifier abundance
   <chr>     <chr>     <chr>          <dbl>
 1 control   1         Q15149          29.9
 2 control   2         Q15149          30.0
 3 control   3         Q15149          29.9
 4 knockdown 1         Q15149          30.4
 5 knockdown 2         Q15149          30.0
 6 knockdown 3         Q15149          30.5
 ...
```

#### The normalization function

For the median normalization method, all samples will be adjusted such that their medians are aligned. Therefor, we need to take the median value of each sample, and the median of the medians, such that they all align.

``` r
normalize_median <- function(
    data = NULL,
    data_centered = NULL
){

  # visible bindings
  abundance <- NULL

  # compute the median shift
  data_centered <- data_centered %>%
    mutate(shift = median(abundance) - abundance)

  # apply the median shift and return the data
  data %>%
    left_join(data_centered,
              by = c('sample','replicate'),
              suffix = c("","_median")) %>%
    mutate(abundance_normalized = abundance + shift) %>%
    select(identifier, sample, replicate, abundance_normalized) %>%
    return()
}
```

#### Normalized data gets merged on return

The data, upon return to the controller function, has the abundance values converted back to non-log values, and merged into the main data object such that all abundance values are avaiable for future selection.

``` r
d_norm <- d_norm %>%
      dplyr::mutate(abundance_normalized = invlog2(abundance_normalized)) %>%
      dplyr::rename(abundance = abundance_normalized)

    data <- data %>% merge_quantitative(d_norm, m)
    
# A tibble: 42,330 × 6
   sample_id sample    replicate protein abundance_raw abundance_median
   <chr>     <chr>     <chr>     <chr>           <dbl>            <dbl>
 1 e9b20ea7  control   1         Q15149    1011259992.       999619474.
 2 ef79cc4c  control   2         Q15149    1093277593.      1106008756.
 3 eebba67b  control   3         Q15149     980809516.      1094113651.
 4 ebf4b0fe  knockdown 1         Q15149    1410445367.      1062933343.
 5 ea36dac9  knockdown 2         Q15149    1072305561.      1113989578.
 6 ecfd1822  knockdown 3         Q15149    1486561518.      1076375880.
 7 e9b20ea7  control   1         Q09666     659299359.       651710227.
 8 ef79cc4c  control   2         Q09666     717783135.       726141683.
 9 eebba67b  control   3         Q09666     612673576.       683450264.
10 ebf4b0fe  knockdown 1         Q09666    1000041657.       753646788.
```

```{r, warning=FALSE, fig.height=3, fig.width=8}
hela_proteins %>% normalize(.method = 'median') %>% plot_normalization()
```

Check out some of the other normalization functions to see how they are implemented.
