---
title: "Subsetting"
output: rmarkdown::html_vignette
description: |
  tidyproteomics: simplified data object and analysis functions
vignette: >
  %\VignetteIndexEntry{Subsetting}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1041)
options(dplyr.print_max = 10)
```

Subsetting data in tidyproteomics with `subset()` is straight forward,
and similar to the tidyverse function `tidyverse::filter()`. The name
`subset()` is used here to avoid conflicting with `dplyr::filter()` and
to also separate it as a tidyproteomics specific function due to the
underlying nature of the tidyproteomics data structure. The function
`subset()` allows data to be easily filtered with simple semantic
expressions, similar to how the filter function in the tidyverse package
dplyr operates. This package also introduces two new operators that work
as a regular expression filter, similar to SQL syntax, (***%like%*** and
***%!like%***) which can be used in the semantic expression to subset
data base on pattern matching in variable groups. For example, the
expression ***`description %!like% 'ribosome'`*** would keep all
proteins with a description that does not include the word *'ribosome'*.
Additionally, together with the `merge()` and `reassign()` functions,
data can be combined from multiple sources, assigned to specific sample
groups and analyzed in a single collective. Alternatively, for example,
data can be separated, normalized and imputed independently then
recombined back into a single collective for analysis and visualization.

```{r, message = FALSE}
library("dplyr")
library("tidyproteomics")

ecoli_proteins %>% summary('sample')
```

### Examples
Subsetting can be a powerful way to slice-and-dice data, print quick
stats or provide a quick visualization. For example, the ecoli data set
can be subset to just the proteins with "Ribosome" in the description:

```{r, message = FALSE}
ecoli_proteins %>% subset(description %like% "Ribosome") %>% summary('sample')
```

Aside from filtering directly on `protein_accession` values, subsetting
can use any of the columns in the `experiments` table:

```{r}
colnames(ecoli_proteins$experiments)
```

any of the terms in the `accounting` table:

```{r}
colnames(ecoli_proteins$accounting)
```

and any of the terms in the `annotations` table:

```{r}
ecoli_proteins$annotations$term %>% unique()
```

### Using Annotations
This allows for the specific importing of specialized terms with
`annotate()` from which subsetting can be performed.

```{r, message = FALSE}
ecoli_proteins %>% subset(cellular_component %like% "nucleus") %>% summary('sample')
```

Additionally, provided the quantitative platform produces an imputed
value commonly referred to as "match between runs", the data can be
filtered to exclude these values. This can be valuable in cases were
true presence/absence is desired, larger portions of the proteome
differ.

### Using Accountings
```{r, message = FALSE}
ecoli_proteins %>% subset(match_between_runs == FALSE) %>% summary('sample')
```

Also, data can be filtered to proteins containing a desired number of
underlying peptides.

```{r, message = FALSE}
ecoli_proteins %>% subset(mum_peptides <= 1) %>% summary('sample')
```

```{r, message = FALSE}
ecoli_proteins %>% subset(num_unique_peptides <= 1) %>% summary('sample')
```

```{r, message = FALSE}
ecoli_proteins %>% 
  subset(cellular_component %like% "cytosol") %>% 
  summary() 
```

### Split then Merge
Here is an example where data is split into two groups, independently
manipulated, the merged back together. Not advisable for an experiment
like this, just for demonstration purposes.

```{r, message = FALSE}
data_ko <- ecoli_proteins %>% 
  subset(sample %like% "ko") %>% 
  normalize(.method = c('median')) %>%
  impute()

data_wt <- ecoli_proteins %>% 
  subset(sample %like% "wt") %>% 
  normalize(.method = c('median')) %>%
  impute()

data_new <- merge(list(data_ko, data_wt), quantitative_source = 'all')
```

```{r, message = FALSE}
data_new
```

```{r, message = FALSE}
data_new %>% summary('sample')
```

```{r}
data_new %>% operations()
```
