---
title: "Summarizing"
output: rmarkdown::html_vignette
description: |
  tidyproteomics: simplified data object and analysis functions
vignette: >
  %\VignetteIndexEntry{Subsetting}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1041)
options(dplyr.print_max = 10)
```

Summarizing proteomics data is vital to understanding the bigger picture
and conveying summary stats that set the tone for the larger analysis.
The results of each summary can be directed to via the destination
option to `"print"` on screen, `"save"` to a file or `"return"` as a
tibble.

```{r, message = FALSE}
library("dplyr")
library("tidyproteomics")
```

### Printing

The tidyproteomics data object can be printed to show a summary of the
object contents.

```{r}
ecoli_proteins
```

And these can expand to encompass additions generated through other
operations.

```{r}
ecoli_proteins %>% expression(ko/wt) %>% enrichment(ko/wt, .term = 'biological_process')
```

### Summarizing

#### Groups

The tidyproteomics data object can summarize the quantitative and
accounting data ...

```{r}
ecoli_proteins %>% summary()
```

... using columns corrisponing to experimental meta-data ...

```{r}
ecoli_proteins %>% summary(by = 'sample')
```

... terms in the annotation meta-data ...

```{r}
ecoli_proteins %>% summary(by = 'biological_process')
```

... and even terms in the accounting data ...

```{r}
ecoli_proteins %>% summary(by = 'num_peptides')
```

```{r}
ecoli_proteins %>% summary(by = 'match_between_runs')
```

#### Contamination

In addition, the data can account for any type of contamination. If the
key word "CRAP" is use then the data will partition out by *Keratin*,
*BSA*, *Trypsin* and *Other*, so long as the protein descriptions
contain CRAP and presumably originate from the
[CRAPome](https://reprint-apms.org/?q=chooseworkflow). However, any
FASTA file, used at the data translation step (eg with
ProteomeDiscoverer), and manipulated to contain CRAP in the description
can be utilized.

```{r}
ecoli_proteins %>% summary(contamination = 'CRAP')
```

Alternatively, any protein descriptor can also be use ...

```{r}
ecoli_proteins %>% summary(contamination = 'Trypsin')
```

... and as such.

```{r}
ecoli_proteins %>% summary(contamination = 'ribosome')
```

### Visualizing

Currently two summary visualizations are implemented in tidyproteomics.

#### Counts

In a simple grouped barchart, proteins (individual, and groups) as well
as peptides (all, unique) are displayed with the match-between-runs
shown as the margin above the ms2 evidenced identifications.

```{r}
ecoli_proteins %>% plot_counts()
```

#### Quantitation

In recent literature a summary of protein quantitation has been
visualized as a rank-based dot-plot.

```{r}
ecoli_proteins %>% plot_quantrank()
```

Additionally, this plot can be extended to highlight statistical
differences via an unbiased all-pair-wise comparison, to give an
anticipated view of how to guide downstream analyses.

```{r}
ecoli_proteins %>% plot_quantrank(display_filter = 'log2_foldchange', display_cutoff = 2)
```

Filter to limit to the range `c(low, high)` and display the protein ids.

```{r}
ecoli_proteins %>% plot_quantrank(show_rank_scale = TRUE, limit_rank = c(1,25))
```
