---
title: "Visualizing"
output: rmarkdown::html_vignette
description: |
  tidyproteomics: simplified data object and analysis functions
vignette: >
  %\VignetteIndexEntry{Subsetting}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1041)
options(dplyr.print_max = 10)
```

### Pre Normalization

Critical to any processing pipeline is the ability to summarize and
visualize data, both pre and post processing. Tidyproteomics covers this
well with both a `summary()` function and several `plot_()` functions.
The summary function (described in more detail
`vignette("summarizing")`) utilizes the same syntax inherent to
`subset()` to generate summary statistics on any variable set, including
all annotated and accounting terms.

### Post Normalization

Visualizing data post processing is an important aspect of data analysis
and great care is taken to explore the data post normalization with a
variety of plot functions. Each of these are intended to display graphs
that should lend insights such as the quantitative dynamic ranges pre
and post normalizations `plot_normalization()`, the sample specific CVs,
dynamic grange `plot_variation_cv()` and principal component variation
`plot_variation_pca()` for each normalization.

```{r, message = FALSE}
library("dplyr")
library("tidyproteomics")
rdata <- ecoli_proteins %>% normalize()

rdata %>% plot_normalization()
```

### Variation

#### Coefficient of Variation and Dynamic Range

The statistical assessment often referred to as CVs ([Coefficient of
Variation](https://en.wikipedia.org/wiki/Coefficient_of_variation)) or
RSD (Relative Standard Deviation) attempts to measure the dispersion in
a measurement. CVs in proteomics is plural because we often measure
hundreds or thousands of proteins simultaneously. Understanding that
variability and the effects of normalization will help improve the
accuracy of your experiments.

```{r}
rdata %>% plot_variation_cv()
```

#### Principal Component Analysis

This is a plot of the accumulative variation explained by each of the
principal components. Ideally, normalization show improve the first few
principal components, removing the measurement and instrument
variability, exposing the underlying biological variability. This plot
show help visuallize that.

```{r}
rdata %>% plot_variation_pca()
```

#### Dynamic Range

Perhaps more intriguing is the plot in `plot_dynamic_range()` which
shows a density heat map of sample specific CVs in relation to
quantitative abundance. This plot highlights how CVs increase at the
lower quantitative range and, more importantly, how each normalization
method can address these large variances. Again, note how random forest
normalization is best able to minimize the CVs at the lower quantitative
range.

```{r}
rdata %>% plot_dynamic_range()
```

### Clustering

Once normalization and imputation methods have been implemented and
selected it is often desired to visualize the unbiased clustering of
samples. This can be accomplished with the `plot_heatmap()` and
`plot_pca()` functions to generate plots.

#### Heatmap

```{r}
rdata %>% plot_heatmap()
```

#### PCA

```{r}
rdata %>% plot_pca()
```

Plot alternative variables, in this case the 3rd and 4th prinicpal
components.

```{r}
rdata %>% plot_pca(variables = c('PC3', 'PC4'))
```
