---
title: "Workflows: Subsetting"
output: rmarkdown::html_vignette
description: |
  r - omics: coded wiring for local analysis of quantitative omics data
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Workflows: Subsetting}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1041)
options(dplyr.print_max = 10)
```

Filtering data with `subset()` allows for the removal or selective
inclusion of subsets without your data, all based on regular expressions
against specific annotations. The most common subset is to remove
contaminant proteins, designated in in our lab with *CRAP* appended to
the beginning of the description.

```{r, message = FALSE}
library("dplyr")
library("tidyproteomics")
```

before

```{r}
ecoli_proteins %>% summary()
```

after removing just the contaminant proteins, here the regex `^` denotes
"it must start with". Learn more about regular expressions here
[www.regular-expressions.info](https://www.regular-expressions.info/).

```{r}
ecoli_proteins %>% 
  subset(description %!like% "^CRAP") %>% 
  summary()
```

In some cases, researchers will co-culture organisms, yet only want to
study and report a specific organism.

```{r, eval = FALSE}
ecoli_proteins %>% subset(description %!like% "Pseudomonas")
```

In other cases, researchers want to study and report on a group of
proteins characterized by a GO term

```{r}
ecoli_proteins %>% 
  subset(cellular_component %like% "cytosol") %>% 
  summary() 
```
