---
title: "Importing Demo"
output: rmarkdown::html_vignette
description: |
  tidyproteomics: simplified data object and analysis functions
vignette: >
  %\VignetteIndexEntry{Importing Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1041)
options(dplyr.print_max = 10)
```

The following is a demonstration workflow for importing data from a public repository. For an additional importing demonstration see `vignette("subsetting")`. All the of examples presented here are from previously published data where the raw data was analyzed on each platform with basic settings for mass tolerance and PTMs.

### ProteomeDiscoverer

Exporting the data requires some setup for the correct columns. This is explained in `vignette("importing")`.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
library(tidyverse)
library(tidyproteomics)

# download the data
url <- "https://data.caltech.edu/records/aevwq-2ps50/files/ProteomeDiscoverer_2.5_p97KD_HCT116_proteins.xlsx?download=1"
download.file(url, destfile = "./data/pd_proteins.xlsx")

# import the data
data_prot <- "./data/pd_proteins.xlsx" %>% import('ProteomeDiscoverer', 'proteins')
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidyproteomics)

# import the data
data_prot <- "~/Local/data/tidyproteomics/ProteomeDiscoverer_2.5/p97KD_HCT116_proteins.xlsx" %>% import('ProteomeDiscoverer', 'proteins')

data_prot
```

### MaxQuant

Exporting the data requires some setup for the correct columns. This is explained in `vignette("importing")`.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
library(tidyverse)
library(tidyproteomics)

# download the data
url <- "https://data.caltech.edu/records/aevwq-2ps50/files/MaxQuant_1.6.10.43_proteinGroups.txt?download=1"
download.file(url, destfile = "./data/mq_proteinGroups.txt")

# import the data
data_prot <- "./data/mq_proteinGroups.txt" %>% 
  import('MaxQuant', 'proteins')
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidyproteomics)

# import the data
data_prot <- "~/Local/data/tidyproteomics/MaxQuant_1.6.10.43/proteinGroups.txt" %>% 
  import('MaxQuant', 'proteins')

data_prot
```

### FragPipe

Exporting the data requires some setup for the correct columns. This is explained in `vignette("importing")`.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
library(tidyverse)
library(tidyproteomics)

# download the data
url <- "https://data.caltech.edu/records/aevwq-2ps50/files/FragPipe_19.1_combined_protein.tsv?download=1"
download.file(url, destfile = "./data/fp_combined_protein.tsv")

# import the data
data_prot <- "./data/fp_combined_protein.tsv" %>% 
  import('FragPipe', 'proteins')
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidyproteomics)

# import the data
data_prot <- "~/Local/data/tidyproteomics/FragPipe_19.1/combined_protein.tsv" %>% 
  import('FragPipe', 'proteins')

data_prot
```
